Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review

### 2017-11-23 analysis/paper/paper.Rmd
grammarly checked

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

# References

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

>>>>

Anaphylaxis is a life-threatening immunologic reaction to an allergen (most frequently food, drugs, insect venom) that is hardly avoidable in clinical practice of various medical specialties. Anaphylaxis is considered a clinical diagnosis and it is highly likely if there is an acute onset of skin symptoms with either hypotension or respiratory compromise, or when the allergen is known or likely [@Sampson2006]. Its pathogenesis relies on a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Anaphylaxis remains a clinical diagnosis [@Kemp2014], but consensus criteria of anaphylaxis suggest a standardized set of criteria [@Sampson2006]. The definition of _refractory_ anaphylaxis is not established. For the purpose of this review, we will define it as anaphylaxis (meeting the criteria by NIAID/FAAN [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

It is important to distinguish refractory anaphylaxis from recurrent anaphylaxis. In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab. We did not include such reports in our analysis and advise on using the term "recurrent" anaphylaxis instead of "refractory" in such cases.

## Epidemiology

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is usually food dependent and less severe than episodes in adulthood. Cases of _refractory_ anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our experience) to be 3-5% of all cases of anaphylaxis.

## Risk factors for anaphylaxis

Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013]. Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigorous exercise (e.g. jogging) [@Worm2013]. Risk factors for _refractory_ anaphylaxis were not investigated previously.

## Management

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. The mainstay of therapy is the early administration of intramuscular (i.m.) adrenalin [@Lazar2017].
However, optimal anaphylaxis treatment is difficult to study because it is an emergency condition and placebo-controlled studies are impossible. Therefore most of the knowledge is based on expert recommendations which are graded as low-level evidence sources.

We aimed to reevaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.

<!-- We compared the successful procedures from the case reports to ??? -->

<!-- ## Eligibility criteria -->
<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale. -->

<!-- ## Information sources -->
<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->

## Literature search strategy

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenalin were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenalin. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included 29 Papers reporting ` r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenalin. Study outcomes were defined as time to elicit a response and mortality. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports therefore the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

<!-- ## Data items -->
<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->

<!-- ## Risk of bias in individual studies -->
<!-- Describe methods used for assessing the risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis. -->

<!-- ## Summary measures -->

<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->

<!-- ## Synthesis of results -->

<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->

<!-- ## Risk of bias across studies  -->

<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->

<!-- ## Additional analyses  -->

<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->

```{r demograph}
source("../../R/data_handling.R")
knitr::kable(demogr.tab,col.names = c("Elicitor","n","Age","Male fraction","Operative fraction"),digits = 1)
```

# Risk factors for refractory anaphylaxis

The risk factors for refractory anaphylaxis include perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the reaction happened during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Surprisingly, no cases of insect venom anaphylaxis were found among these reports.cWhile the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities. [@Nag2017]

<!-- # Firs sings in anesthesia setting of anaphylaxis -->
Early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.

# Therapy of refractory anaphylaxis

Most of the treatment strategies were according to the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed. We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.

Based on the analyzed reports, ambulatory patients were first treated with adrenaline were patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, followed by steroid and antihistaminic agents. Epinephrine was given in 100% of cases in various doses.

<!-- In Table X we provided therapeutic approaches used in the treatment of anaphylaxis.  -->

4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.

```{r tabFatal}
rownames(fat) <- NULL
colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
knitr::kable(fat)
```

World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011]. Patients should be transported to a critical care unit. AAAAI [@Lieberman] suggest similar treatment but include dopamine, vasopressin if epinephrine injections and volume expansion fail to alleviate hypotension, glucagon atropine, and methylene blue. EAACI [@Muraro2014] guidelines also suggest using glucagon in case of concomitant beta-blockade.

# Methylene blue for the treatment of refractory hypotension

`r which(d$What.helped_gr=="methylene blue") %>% length` reports (`r ((which(d$What.helped_gr=="methylene blue") %>% length)/length(d$What.helped_gr)*100) %>% round(1)`%) reported using methylene blue to treat refractory anaphylaxis and indicated a rapid (within minutes) increase in blood pressure after the injection of methylene blue. It was both given as a bolus and as an infusion with the doses ranging from 0.5-2mg/kg. Patients responded well to the treatment with methylene blue and a second infusion (or bolus) were successfully given in case of a recurrent hypotension within hours of the anaphylactic episode.

The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce Nitric oxide (NO). NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000]. This in effect leads to the opening of the calcium ion channels in smooth muscle tissues and subsequent loss of vascular muscle contractility [@Omar2015].

Methylene blue is a well-investigated drug used for treating patients with methemoglobinemia [@Schirmer2003]. Buzato et al. investigated its properties as an inhibitor of nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue was mediated by the sGC inhibition. Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].

Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013]. Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) *[7-9]* production which is a potent mediator of anaphylaxis [].

The use of methylene blue in current management of anaphylaxis is scarce. We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.

We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension using 100 mg bolus i.v. and follow it with an infusion of 50 µg/kg/min. As a vector, a 100 mL of 5% dextrose can be used.

As methylene blue does not block nitric oxide production its side effects are minimal (Nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis. No side effects were reported due to methylene blue use in the above cases.

It is important to remember that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].

# Vasopressors Vasopressin as a supplementary vasopressor

In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated.

Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol, methoxamine, dopamine and noradrenaline infusions.

Our register lists 18 cases where dopamine was used. None were fatal.

Use of vasopressin in refractory anaphylaxis was only suggested in the AAAAI management criteria [@Liberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine[@Avni]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@Chiara2008].

Seedat et al. described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia. The refractory shock was not responding to repeated adrenalin (total 2mg), antihistaminics and hydrocortisone. Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when biphasic reaction occurred. Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued. [@Seedat2014] We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.

One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell]. Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.

# Glucagon in anaphylaxis with concomitant beta blockers

Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986, @Javeed1996].  Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986], Glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers.

Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis. [@Nag2017]*

# ECLS

`r (d$What.helped_gr=="ECLS") %>% which()`
*The incidence of life-threatening anaphylactic reactions related to anesthesia is approximately 1 in 6000 anesthetics administered and is associated with mortality as high as 5%.1 The use of extracorporeal membrane oxygenation (ECMO) in the setting of refractory shock following anaphylaxis has been documented.2-5 [@Weiss2015]*

*The use of VA ECMO for treating cardiogenic shock is well established, as are the many pitfalls and complica- tions associated with its use.10 VA ECMO may also be used for treating anaphylactic shock,2-5 although its role is less well established than for treating cardiogenic shock of other etiology [@Weiss2015]*

*Patient with refractory anaphylaxis to Chlorhexidine coated CVC. 10 minutes after introducing CVC he became hypotensive - then >30min resuscitation and 3 minutes after CVC removal - the return of spontaneous circulation. [@Wang2016] This case shows how important it is to get rid of the allergen in refractory anaphylaxis. Treatment with ECMO.*

# Stopping the elicitor exposure
In cases where the elicitor of anaphylaxis is known, and first-line  therapy fails it is advisable to stop the exposition to the likely allergen. Three case reports described the rapid improvement of a refractory anaphylaxis after preventing further contact with the allergen.

A 15-year-old boy who was exposed to cow's milk underwent near-fatal anaphylaxis with refractory hypotension. After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylaxis symptoms and patient recovered fully. Authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].

Another report used sugammadex which is a compound strongly binding to rocuronium, to decrease its bioavailability in a patient under anesthesia and therefore allow for a cessation of a refractory anaphylaxis episode [@Raft2012].

There also was a report where the patient was inadvertently given i.v. high dose of aprotinin in a fast infusion. After 10 hours of refractory hypotension, authors decided to use the last resort treatment with high-volume continuous venovenous hemofiltration (HV-CVVH) which resulted in a rapid improvement of hemodynamic function [@Lango2009].

# Conclusion 
<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article. Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->

# Expert Commentary: 500-1000 words (included in overall word count).
<!-- **To distinguish the articles published in the Expert Review series, authors must provide an additional section entitled ‘Expert Commentary’. This section affords authors the opportunity to provide their interpretation of the data presented in the article and discuss the developments that are likely to be important in the future, and the avenues of research likely to become exciting as further studies yield more detailed results. The intention is to go beyond a conclusion and should not simply summarise the paper. Authors should answer the following:** -->

## What are the key weaknesses in clinical management so far?

## What potential does further research hold? What is the ultimate goal in this field?

## What research or knowledge is needed to achieve this goal and what is the biggest challenge in this goal being achieved?

## Is there any particular area of the research you are finding of interest at present?
<!-- **Please note that ‘opinions’ are encouraged in the Expert commentary section, and, as such, referees are asked to keep this in mind when peer reviewing the manuscript.** -->

# Five-year view
<!-- **Authors are challenged to include a speculative viewpoint on how the field will have evolved five years from the point at which the review was written.** -->

# Key issues
<!-- **An executive summary of the authors’ main points (bulleted) is very useful for time-constrained readers requiring a rapidly accessible overview.** -->

<!-- * Matrix-Assisted Laser Desorption/Ionization  Mass Spectrometry Imaging (MALDI-MSI) is able to provide a molecular insight of the samples, detecting the presence of a great variety of analytes directly on-tissue and showing their spatial distribution across the tissue section. -->

*? Epinephrine in appropriate doses is safe, and there are no abso- lute contraindications for its use in treating anaphylaxis. ? Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.*[@Fineman2015]
* Glucagon
* Vasopressors
* ECLS
*

# References
<div id="refs"></div>
<!-- **A maximum of 100 references is suggested. Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography. Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->

## Annotated bibliography
<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->

# Figures and Tables
<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->

```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
source("../../R/prisma.R")
knitr::include_graphics("../figures/prism_flow.png")
```

<!-- ### Colophon -->

<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->

```{r colophon, cache = FALSE}
# which R packages and versions?
# devtools::session_info()
```

<!-- The current Git commit details are: -->

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
# git2r::repository("../..")
```
